Modeling the Java Card API (or Interface?) in B

### 2012-03-28 trunk/papers/B2012-work/javacardAPIpaper.tex
um pouco mais sobre a modelagem.

Among the different components of the Java Card platform, the Java
Card API is the one without which no refinement into Java Card code
could be done.  Suppose, for instance, that we had the B abstract
machine of Fig.~\ref{fig:Bmachine}, which models a bus card with a
variable that keeps track of the current credit the user has (\texttt{balance}) and an operation to add credits to this
balance (\texttt{add\_credit}!!! verificar). This simple example provides enough evidence of the
importance of having a B model of the Java Card API.  As all data in
it is implementable in C, for instance, and it is a deterministic
specification, B tools can directly generate code for it in C, and the
resulting code would be very similar to the original B machine.

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{figs/machineB.png}
\caption{Example B machine.}
\label{fig:Bmachine}
\end{figure}

However, Java Card code corresponding to it would look like the one
presented in Fig.~\ref{fig:jcapplet}. In a rigorous development
process, we do not want the code generator to carry out all these
non-verified transformations in the code. So, we need to be able to
refine the original B machine into something such as the B refinement
of Fig.~\ref{B ref usando a API}. Then, code generation would be
straightforward (as it should be). This is then the reason why it
needed to be the first Java Card component to be modeled in B. (AMM:
maybe also cite related work ? Which are the formalization works that
did not focus on the API?)

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{figs/appletJC.png}
\caption{Example Java Card applet.}
\label{fig:jcapplet}
\end{figure}

COMENT SIMONE -> coloquei duas figuras, veja o que acha do exemplo, é o mesmo que foi usado na dissertação, ele é simples e pequeno. Coloquei a máquina abstrata e o modelo do applet Java Card (veja o que acha do tamanho das figuras), estou vendo com o Bruno como a implementação desta máquina fica com a API, porque o modelo dela que fizemos foi sem o uso da API, ela possui apenas o uso dos tipos primitivos e as operações de soma e comparação deles. O exemplo que nós usamos na validação do trabalho que tinha o uso da API foi um refinamento da máquina applet.

[[[ PARA SIMONE -> Por favor, coloque os exemplos sob a forma de listing. Fica mais facil de editar, controlar o tamanho e conteudo e fica mais legivel.
Quanto ao codigo javacard, basta o da operacao addcredit. Alias, o nome não esta batendo. Nao verifiquei o resto. Por favor verifique. ]]]

[[[ PARA SIMONE -> tambem precisamos com urgencia daquela anelise sobre o que muda da 2.2.2 para a 3 ]]]

@@

@@

>>>>

Among the different components of the Java Card platform, the Java
Card API is the one without which no refinement into Java Card code
could be done.  Suppose, for instance, that we had the B abstract
machine of Fig.~\ref{lis:buscard}, which models a bus card with a
variable that keeps track of the current credit the user has (\texttt{balance}) and an operation to add credits to this
balance (\texttt{add}). This simple example provides enough evidence of the
importance of having a B model of the Java Card API.  As all data in
it is implementable in C, for instance, and it is a deterministic
specification, B tools can directly generate code for it in C, and the
resulting code would be very similar to the original B machine.

\lstset{language=B, caption={B machine Buscard},label=lis:buscard}
\begin{lstlisting}
MACHINE
	Buscard
CONSTANTS
	maximum
PROPERTIES
	maximum : INTEGER & maximum = 100
VARIABLES
	balance
INVARIANT
	balance : INTEGER & balance <= maximum
INITIALISATION
	balance := 0
OPERATIONS
	add(credit) =
	PRE
		credit : INTEGER & credit > 0  & balance + credit <= maximum
	THEN
		balance := balance + credit
	END
END
\end{lstlisting}

However, Java Card code corresponding to the operation \texttt{add} of this B specification would look like the one
presented in Fig.~\ref{lis:addjavacard}. In a rigorous development
process, we do not want the code generator to carry out all these
non-verified transformations in the code. So, we need to be able to
refine the original B machine into something such as the B refinement
of Fig.~\ref{lis:addimplementation}. Then, code generation would be
straightforward (as it should be).

\lstset{language=Java, caption={Method add in applet Java Card},label=lis:addjavacard}
\begin{lstlisting}
public void add(APDU apdu) {
	byte[] buffer = apdu.getBuffer();
	byte credit = (byte) Util.makeShort((byte) buffer[ISO7816.OFFSET_P1], (byte)buffer[ISO7816.OFFSET_P2]);
	if (!(credit >= 0)) {
		ISOException.throwIt(EXCEPTIONS.negative_credit);
	}
	else if (!(balance + credit <= MAXIMUM)) {
		ISOException.throwIt(EXCEPTIONS.balance_exceeded);
	}
	else {
		balance = (balance + credit);
	}
}
\end{lstlisting}

\lstset{language=B, caption={B machine Buscard},label=lis:addimplementation}
\begin{lstlisting}
	add(apdu) =
	BEGIN
		buffer <- - getBuffer();
		credit <-- cast ( );
		IF 
			not(credit >= 0)      !!! byte version
		THEN
			ISOException.throwIt(EXCEPTIONS.negative_credit)
		ELSIF 
			not(balance + credit <= MAXIMUM)     !!! byte version
		THEN
			ISOException.throwIt(EXCEPTIONS.balance_exceeded)
		ELSE
		        balance := sum_byte (balance, credit)  
		END
	END

	[[[[BRUNO e SIMONE: isso esta todo errado - coloquei aqui mais ou menos o que eu acho que deve ser a cara da implementacao B que gera o codigo java card. Voces devem ter algum exemplo para recuperar a estrutura certa, nao?]]]]

\end{lstlisting}

We can identify two main kinds of transformation that had to be carried out in order to obtain the code  of operation \texttt{add} in figs. \ref{lis:addjavacard} and \ref{lis:addimplementation} from the abstract specification in fig. \ref{lis:buscard}:

\begin{enumerate}
\item the transition into a defensive style of programming by the inclusion of explicit tests to check that the input conditions (pre conditions in the  B specification) are satisfied before executing the corresponding action;

\item the inclusion of Java Card specific aspects, such as the change of interface of \texttt{add} due to the need of passing data via the APDU protocol and the manipulation of the APDU buffer by methods offered the Java Card API.

\end{enumerate}

The transition into a defensive style of programming is part of the BSmart method, and implemented by the BSmart Full Function component. This component generates a B refinement of the original specification where pre-conditions are transformed into conditionals (IF-THEN-ELSE-END constructs), with the indication of the need to throw the corresponding exception when conditions are not met by input data and states.
The resulting B model can then be formally verified as usual in the B method.

The change of interface is also taken care of by the BSmart method, so that the development can be done without need to extend the B method. More details on this [[[???? (we haven't published this part but there is the technical report )]]].

But all of these transformations rely, at the bottom, on the need to reference Java Card API types, classes and methods on the B level.
This is then the reason why the Java Card API needed to be the first Java Card component to be modeled in B.

[[[(PARA SIMONE e BRUNO: maybe also cite related work ? Which are the formalization works that did not focus on the API?)]]]

[[[ PARA SIMONE -> tambem precisamos com urgencia daquela analise sobre o que muda da 2.2.2 para a 3 ]]]

@@

\item Each machine corresponding to a Java Card class defines an abstract set with the same name (distinction on character cases, only). For instance, the machine Throwable declares the abstract set THROWABLE. This set can then be used to reference the type corresponding to the class.

\item The hierarchy of the Java Card API objects is partially modeled by abstract set inclusion. For instance, the class Throwable extends the class Object, in Java Card. In our model, we also have that the machine Throwable EXTENDS the machine Object, but the meaning of a machine extension in B is different from the meaning of a class extension in Java Card. To show that all throwables are indeed objects, a property stating that  THROWABLE is a subset of OBJECT is included in the Throwable machine.

@@

[[[ incluir aqui parte da discussao sobre as opcoes de modelagem  com o argumento pela escolha dos subsets para a tipagem ]]] 
As our focus was on the interfaces of the Java Card API methods,
